Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-07-30 analysis/paper/paper.qmd
minor edits

# Kiwulan site and previous studies
This study focuses on archaeological evidence from Kiwulan, a large multi-component site of with an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. Kiwulan was inhabited by Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

The excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, post-holes, and burials. These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. In addition to locally made pottery, porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials. Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area.

>>>>

# Kiwulan site and previous studies
This study focuses on archaeological evidence from Kiwulan, a large multi-component site where an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. Kiwulan was inhabited by the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in the historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, burials, and post-holes with posts in situ. These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. In addition to locally made pottery with geometric patterns, foreign goods such as porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials. Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area, mostly with the Basay people active in northern Taiwan.
